The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-07 script/de/chapters/aboutcryptool.tex
Language de: Use the book class for the script, instead of the article class
CT script: Some small updates and corrections done in German and English

% --------------------------------------------------------------------------
\addcontentsline{toc}{section}{Überblick}  % Hier steht im Inhaltsverzeichnis
                    % die richtige Seitenzahl und beim Klicken springt es an
                    % die richtige Seite; aber alles was nach dem 
                    % Inhaltsverzeichnis hat ein Problem! be_2006 
\section*{Überblick über den Inhalt des CrypTool-Skripts}

\parskip 4pt
%\vskip +12 pt
{
In diesem {\em Skript zu dem Programm CrypTool} \index{CrypTool} finden Sie
eher mathematisch orientierte Informationen zum Einsatz von
kryptographischen Verfahren. Die Hauptkapitel sind von
verschiedenen {\bf Autoren} verfasst (siehe Anhang \ref{s:appendix-authors})
 %\hyperlink{appendix-authors}{Autoren}
und in sich abgeschlossen. Am Ende der meisten Kapitel finden Sie
jeweils Literaturangaben und Web-Links.

@@

Das \hyperlink{Chapter_Crypto2020}{letzte Kapitel} {\bf Krypto2020}
diskutiert die Gefahren für bestehende kryptographische Verfahren und 
stellt aktuelle, alternative Forschungsansätze für eine langfristige
kryptografische Sicherheit vor.

@@

\vskip +7 pt \noindent
Die aktuelle Version von CrypTool\index{CrypTool} finden Sie jeweils
auf den Webseiten 
\newline
  \href{http://www.CrypTool.de}{\texttt{http://www.CrypTool.de}}, ~
  \href{http://www.cryptool.com}{\texttt{http://www.CrypTool.com}} ~ oder  ~
  \href{http://www.cryptool.org}{\texttt{http://www.CrypTool.org}}.
\vskip + 7 pt \noindent
Die Ansprechpartner für dieses kostenlose Open Source-Programm\index{Open
Source} sind auf der Webseite und in der zum CrypTool-Paket dazugehörenden
Readme-Datei genannt.
}

>>>>

% --------------------------------------------------------------------------
\addcontentsline{toc}{chapter}{Überblick}  % Hier steht im Inhaltsverzeichnis
                    % die richtige Seitenzahl und beim Klicken springt es an
                    % die richtige Seite; aber alles was nach dem 
                    % Inhaltsverzeichnis hat ein Problem! be_2006 
\chapter*{Überblick über den Inhalt des CrypTool-Skripts}

\parskip 4pt
%\vskip +12 pt
{
In diesem {\em Skript zu dem Programm CrypTool} \index{CrypTool} finden Sie
eher mathematisch orientierte Informationen zum Einsatz von
kryptographischen Verfahren. Zu einigen Verfahren gibt es Beispielcode,
geschrieben für das CAS (Computer-Algebra-System) Sage.
Die Hauptkapitel sind von verschiedenen {\bf Autoren} verfasst
(siehe Anhang \ref{s:appendix-authors}) %\hyperlink{appendix-authors}{Autoren}
und in sich abgeschlossen. Am Ende der meisten Kapitel finden Sie
jeweils Literaturangaben und Web-Links.

@@

Das \hyperlink{Chapter_Crypto2020}{letzte Kapitel} {\bf Krypto 2020}
diskutiert Bedrohungen für bestehende kryptographische Verfahren und 
stellt alternative Forschungsansätze für eine langfristige
kryptografische Sicherheit vor.

@@

### 2009-07-07 script/de/chapters/authors.tex
Language de: Use the book class for the script, instead of the article class
CT script: Some small updates and corrections done in German and English

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{appendix-authors}{}
\subsection{Autoren des CrypTool-Skripts\index{CrypTool}}
\label{s:appendix-authors}

@@

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{appendix-movies}{}
\subsection{Filme und belletristische Literatur mit Bezug zur Kryptographie,\\
Bücher mit Sammlungen einfacher Verfahren für Kinder}
\label{s:appendix-movies}  
% {\bf Filme und Literatur mit Bezug zur Kryptographie} (siehe Anhang \ref{s:appendix-movies})
\index{Filme}
\index{Literatur}

@@

\subsubsection{Für Erwachsene}

@@

\vskip +60 pt
\subsubsection{Für Kinder}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{appendix-Learn-NT}{}
\subsection{Lernprogramm Elementare Zahlentheorie}
    \label{s:appendix-Learn-NT}  
    \index{ZT, Lernprogramm Zahlentheorie}%
    \index{Lernprogramm ZT}%

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\clearpage   % Nötig, damit Lit.verzeichnis nicht gl. nach Abb. 8 kommt.
% \newpage
\begin{thebibliography}{99999}
%\addcontentsline{toc}{subsection}{Literaturverzeichnis}

>>>>

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{appendix-authors}{}
\section{Autoren des CrypTool-Skripts\index{CrypTool}}
\label{s:appendix-authors}

@@

\item[Minh Van Nguyen,] \mbox{}\\
Sage-Entwickler und Reviewer des Dokuments.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{appendix-movies}{}
\section{Filme und belletristische Literatur mit Bezug zur Kryptographie,\\
Bücher mit Sammlungen einfacher Verfahren für Kinder}
\label{s:appendix-movies}  
% {\bf Filme und Literatur mit Bezug zur Kryptographie} (siehe Anhang \ref{s:appendix-movies})
\index{Filme}
\index{Literatur}

@@

\subsection{Für Erwachsene}

@@

\vskip +60 pt
\subsection{Für Kinder}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{appendix-Learn-NT}{}
\section{Lernprogramm Elementare Zahlentheorie}
    \label{s:appendix-Learn-NT}  
    \index{ZT, Lernprogramm Zahlentheorie}%
    \index{Lernprogramm ZT}%

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\clearpage   % Nötig, damit Lit.verzeichnis nicht gl. nach Abb. 8 kommt.
% \newpage
\begin{thebibliography}{99999}
%\addcontentsline{toc}{section}{Literaturverzeichnis}

### 2009-07-07 script/de/chapters/ellipticcurves.tex
CT script: Some small updates and corrections done in German and English

\begin{enumerate}
  \item Online-Tutorial über Elliptische Kurven der Firma Certicom\index{Certicom}\\
        \href{http://www.certicom.com/resources/ecc_tutorial/ecc_tutorial.html}
	{\texttt{http://www.certicom.com/resources/ecc\_tutorial/ecc\_tutorial.html}}
  \item Arbeitsgruppe IEEE P1363 \\
        \href{http://grouper.ieee.org/groups/1363}
	{\texttt{http://grouper.ieee.org/groups/1363}}
  \item \hypertarget{Lenstra2} Eine informative Seite zum Faktorisieren mit Elliptischen Kurven. \\
        \href{http://www.loria.fr/~zimmerma/records/ecmnet.html}{\texttt{http://www.loria.fr/\~{}zimmerma/records/ecmnet.html}} \\
        Dort findet man Literatur zum Thema Faktorisieren mit Elliptischen
	Kurven sowie Links zu anderen ECC-Seiten.
  \item Schlüssellängen-Vergleich von Arjen Lenstra und Eric Verheul\\
        \href{http://cryptosavvy.com/table.htm}
	{\texttt{http://cryptosavvy.com/table.htm}}
\end{enumerate}

>>>>

\begin{enumerate}
   \item Umfassende interaktive Einführung in elliptische Kurven und
         Elliptische-Kurven-Kryptographie (ECC) mit Sage\index{Sage}
         von Maike Massierer und dem CrypTool-Team (alles in Englisch),\\
         \href{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}
      {\texttt{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}}

  \item Online-Tutorial über Elliptische Kurven der Firma Certicom\index{Certicom},\
        \href{http://www.certicom.com/resources/ecc_tutorial/ecc_tutorial.html}
     {\texttt{http://www.certicom.com/resources/ecc\_tutorial/ecc\_tutorial.html}}

  \item Arbeitsgruppe IEEE P1363, \\
        \href{http://grouper.ieee.org/groups/1363}
     {\texttt{http://grouper.ieee.org/groups/1363}}

  \item \hypertarget{Lenstra2} Eine informative Seite zum Faktorisieren mit
        Elliptischen Kurven, \\
        \href{http://www.loria.fr/~zimmerma/records/ecmnet.html}
     {\texttt{http://www.loria.fr/\~{}zimmerma/records/ecmnet.html}} \\
        Dort findet man Literatur zum Thema Faktorisieren mit Elliptischen
	Kurven sowie Links zu anderen ECC-Seiten.

  \item Schlüssellängen-Vergleich von Arjen Lenstra und Eric Verheul,\\
        \href{http://cryptosavvy.com/table.htm}
	{\texttt{http://cryptosavvy.com/table.htm}}
\end{enumerate}

### 2009-07-07 script/de/chapters/introduction.tex
CT script: Some small updates and corrections done in German and English

Neu hinzu gekommen ist der Anhang über die Benutzung des Computer-Algebra-Systems
Sage \ref{s:appendix-using-sage}. Dieses wurde ergänzt, weil Sage mehr und mehr
zum Standard-Open-Source CAS wird. Entsprechend wurden alle Codebeispiele, die
vorher in PARI-GP und Mathematica geschrieben waren, durch Sage-Code ersetzt.
Außerdem kamen dank Minh Van Nguyen etliche neue Code-Beispiele hinzu.

@@

Ich hoffe, dass viele Leser mit diesem Skript mehr Interesse an und 
Verständnis für dieses moderne und zugleich uralte Thema finden.\\[1.5\baselineskip]
\enlargethispage*{2\baselineskip}
\nopagebreak
Bernhard Esslinger\\[\baselineskip]
\nopagebreak
Frankfurt, July 2009
\vspace{-12pt}

@@

An dieser Stelle möchte ich explizit folgenden Personen danken, 
die bisher in ganz besonderer Weise zu CrypTool\index{CrypTool} 
beigetragen haben.
Ohne ihre besonderen Fähigkeiten und ihr großes Engagement wäre CrypTool
nicht, was es heute ist:
\vspace{-7pt}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Hr.\ Henrik Koy
   \item Hr.\ Jörg-Cornelius Schneider
   \item Hr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Mitarbeiter von Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann und Prof.\ Torben Weis.
\end{list}
Auch allen hier nicht namentlich Genannten ganz herzlichen Dank für das 
(meist in der Freizeit) geleistete Engagement.
\\

Bernhard Esslinger

Frankfurt, Juli 2009

>>>>

Neu hinzu gekommen ist der Anhang über die Benutzung des Computer-Algebra-Systems
Sage (siehe \ref{s:appendix-using-sage}). Dieses wurde ergänzt, weil Sage mehr und mehr
zum Standard-Open-Source CAS wird. Entsprechend wurden alle Codebeispiele, die
vorher in PARI-GP und Mathematica geschrieben waren, durch Sage-Code ersetzt.
Außerdem kamen dank Minh Van Nguyen etliche neue Code-Beispiele hinzu.

@@

Ich hoffe, dass viele Leser mit diesem Skript mehr Interesse an und 
Verständnis für dieses moderne und zugleich uralte Thema finden.
\\
\\
% [1.5\baselineskip]
% \enlargethispage*{2\baselineskip}
% \nopagebreak
Bernhard Esslinger
\\
\\
% [\baselineskip]
% \nopagebreak
Frankfurt, Juli 2009
\vspace{-12pt}

@@

An dieser Stelle möchte ich explizit folgenden Personen danken, 
die bisher in ganz besonderer Weise zu CrypTool\index{CrypTool} 
beigetragen haben.
Ohne ihre besonderen Fähigkeiten und ihr großes Engagement wäre CrypTool
nicht, was es heute ist:
\vspace{-7pt}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Hr.\ Henrik Koy
   \item Hr.\ Jörg-Cornelius Schneider
   \item Hr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Mitarbeiter von Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann und Prof.\ Torben Weis.
\end{list}
Auch allen hier nicht namentlich Genannten ganz herzlichen Dank für das 
(meist in der Freizeit) geleistete Engagement.
\\
\\
Bernhard Esslinger
\\
\\
Frankfurt, Juli 2009

### 2009-07-07 script/de/chapters/numbertheory.tex
CT script: Some small updates and corrections done in German and English

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{RSAKonkret}{}%
\hypertarget{Chapter_ElementaryNT_12}{}%ist wahrscheinlich redundant !
\section{Das RSA-Verfahren mit konkreten Zahlen}
\label{rsakonkret}\index{RSA}
Nachdem oben die \hyperlink{RSA}{Funktionsweise des RSA-Verfahrens} beschrieben
wurde, sollen diese Schritte hier mit konkreten, aber kleinen Zahlen
durchgeführt werden.

@@

\bibitem[Agrawal2002]{Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002 \\
    \href{http://www.cse.iitk.ac.in/news/primality.html}
    {\texttt{http://www.cse.iitk.ac.in/news/primality.html}}

>>>>

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{RSAKonkret}{}%
\hypertarget{Chapter_ElementaryNT_12}{}%ist wahrscheinlich redundant !
\section{Das RSA-Verfahren mit konkreten Zahlen\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Minh Van Nguyen}%
    Weiteres Material: Minh Van Nguyen: \glqq Number Theory and the RSA Public Key
    Cryptosystem. Introductory tutorial on using Sage to study elementary number theory
    and public key cryptography'', 2009.
    Didaktisch sehr klarer Artikel zu einigen Grundlagen der Zahlentheorie und zur
    Benutzung von Sage.
    \href{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}
       {\texttt{http://nguyenminh2.googlepages.com/sage\_numtheory-rsa.pdf}}.
    \vskip +3 pt
}
\label{rsakonkret}\index{RSA}
Nachdem oben die \hyperlink{RSA}{Funktionsweise des RSA-Verfahrens} beschrieben
wurde, sollen diese Schritte hier mit konkreten, aber kleinen Zahlen
durchgeführt werden.

@@

\bibitem[Agrawal2002]{Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002, Korrigierte Fassung: \\
       \href{http://www.cse.iitk.ac.in/~manindra/algebra/primality_v6.pdf}
    {\texttt{http://www.cse.iitk.ac.in/~manindra/algebra/primality\_v6.pdf}} \\
    Siehe auch die Seite \glqq The AKS \glqq PRIMES in P'' Algorithm Resource'':
       \href{http://fatphil.org/maths/AKS/}
    {\texttt{http://fatphil.org/maths/AKS/}}

### 2009-07-07 script/en/chapters/aboutcryptool.tex
CT script: Some small updates and corrections done in German and English

\parskip 4pt
%\vskip +12 pt
{
\noindent In this CrypTool\index{CrypTool} script you will find predominantly
mathematically oriented information on using cryptographic procedures.
The main chapters have been written by various
{\bf authors} (see appendix \ref{s:appendix-authors})
%\hyperlink{appendix-authors}{authors}
and are therefore independent from one another. At the end of most chapters
you will find literature and web links.

@@

The \hyperlink{Chapter_Crypto2020}{last chapter} {\bf Crypto2020}
discusses the threats for existing cryptographic methods and introduces
the current alternative research approaches to achieve long-term security
of cryptographic schemes.

Whereas the \textit{eLearning program} CrypTool\index{CrypTool} motivates and
teaches you how to use cryptography in practice, the \textit{script} provides
those interested in the subject with a deeper understanding of the mathematical
algorithms used -- trying to do it in an instructive way.
If you are already a little bit familiar with this field of knowledge you can
gain a fast overview about the functions delivered by CrypTool using the
{\bf menu tree} (see appendix \ref{s:appendix-menutree}).

@@

\vskip +7 pt \noindent
You will find the current version of CrypTool\index{CrypTool}
at the web sites \newline
  \href{http://www.cryptool.org}{\texttt{http://www.cryptool.org}},~
  \href{http://www.cryptool.com}{\texttt{http://www.cryptool.com}}~ or~
  \href{http://www.CrypTool.de}{\texttt{http://www.cryptool.de}}.

\vskip +7 pt \noindent
The contact people for this free open-source program are listed at the 
web site and in the ``readme'' file delivered within the 
CrypTool\index{CrypTool} package.}

>>>>

\parskip 4pt
%\vskip +12 pt
{
\noindent In this CrypTool\index{CrypTool} script you will find predominantly
mathematically oriented information on using cryptographic procedures. There is
also sample code in the CAS (computer algebra system) Sage for some of the
procedures.
The main chapters have been written by various {\bf authors}
(see appendix \ref{s:appendix-authors}) %\hyperlink{appendix-authors}{authors}
and are therefore independent from one another. At the end of most chapters
you will find references and web links.

@@

The \hyperlink{Chapter_Crypto2020}{last chapter} {\bf Crypto2020}
discusses threats for existing cryptographic methods and introduces
the alternative research approaches to achieve long-term security
of cryptographic schemes.

Whereas the \textit{eLearning program} CrypTool\index{CrypTool} motivates and
teaches you how to use cryptography in practice, the \textit{script} provides
those interested in the subject with a deeper understanding of the mathematical
algorithms used -- trying to do it in an instructive way.
If you are already a little bit familiar with this field of knowledge you can
gain a fast overview about the functions delivered by CrypTool looking at the
{\bf menu tree} (see appendix \ref{s:appendix-menutree}).

@@

### 2009-07-07 script/en/chapters/ellipticcurves.tex
CT script: Some small updates and corrections done in German and English

\begin{enumerate}
   \item Certicom Online Tutorial, \index{Certicom}\\
         \href{http://www.certicom.com/resources/ecc_tutorial/ecc_tutorial.html}{\texttt{http://www.certicom.com/resources/ecc\_tutorial/ecc\_tutorial.html}}

   \item Working group IEEE P1363 \\
         \href{http://grouper.ieee.org/groups/1363}{\texttt{http://grouper.ieee.org/groups/1363}}

   \item \hypertarget{Lenstra2}{}
         An informative web page about factorisation with elliptic curves. \\
         \href{http://www.loria.fr/~zimmerma/records/ecmnet.html}
	 {\texttt{http://www.loria.fr/\~{}zimmerma/records/ecmnet.html}} \\
         It contains literature related to the topic factorisation with 
	 elliptic curves as well as links to other web page.

   \item Key length comparison by Arjen Lenstra and Eric Verheul\\
         \href{http://cryptosavvy.com/table.htm}
	  {\texttt{http://cryptosavvy.com/table.htm}}

>>>>

\begin{enumerate}
   \item Interactive introduction to elliptic curves and elliptic curve cryptography
         with Sage\index{Sage} by Maike Massierer and the CrypTool team,\\
         \href{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}
      {\texttt{http://sage.mathematik.uni-siegen.de:8000/home/pub/45/}}

   \item Certicom Online Tutorial, \index{Certicom},\\
         \href{http://www.certicom.com/resources/ecc_tutorial/ecc_tutorial.html}
      {\texttt{http://www.certicom.com/resources/ecc\_tutorial/ecc\_tutorial.html}}

   \item Working group IEEE P1363, \\
         \href{http://grouper.ieee.org/groups/1363}
      {\texttt{http://grouper.ieee.org/groups/1363}}

   \item \hypertarget{Lenstra2}{}
         An informative web page about factorisation with elliptic curves, \\
         \href{http://www.loria.fr/~zimmerma/records/ecmnet.html}
	 {\texttt{http://www.loria.fr/\~{}zimmerma/records/ecmnet.html}} \\
         It contains literature related to the topic factorisation with 
	 elliptic curves as well as links to other web page.

   \item Key length comparison by Arjen Lenstra and Eric Verheul,\\
         \href{http://cryptosavvy.com/table.htm}
      {\texttt{http://cryptosavvy.com/table.htm}}

### 2009-07-07 script/en/chapters/introduction.tex
CT script: Some small updates and corrections done in German and English

Starting in the year 2000 this script became a part of the 
CrypTool\index{CrypTool} package. It is designed to accompany the program 
CrypTool by explaining some mathematical topics in more detail, 
but still in a way which is easy to understand.

@@

Newly added is the appendix about using the computer algebra system
Sage \ref{s:appendix-using-sage}. This was added because Sage becomes
more and more the standard open-source CAS system. Accordingly all
samples written before in PARI-GP and Mathematica have been substituted
with Sage code. Thanks to Minh Van Nguyen a lot of new code samples could be added.

@@

I hope that many readers have fun with this script and that they get 
out of it more interest and greater understanding of this modern but 
also very ancient topic.
\\[1.5\baselineskip]
\enlargethispage*{2\baselineskip}
\nopagebreak
Bernhard Esslinger
\\[\baselineskip]
Frankfurt (Germany), July 2009
\vspace{-12pt}

@@

At this point I'd like to thank explicitly the following people who
particularly contributed to CrypTool\index{CrypTool}. They applied their
very special talents and showed really great engagement:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Staff of Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann and Prof.\ Torben Weis.
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

Bernhard Esslinger

Frankfurt (Germany), July 2009

>>>>

Starting in the year 2000 this script became part of the 
CrypTool\index{CrypTool} package. It is designed to accompany the program 
CrypTool by explaining some mathematical topics in more detail, 
but still in a way which is easy to understand.

@@

Newly added is the appendix about using the computer algebra system
Sage (see \ref{s:appendix-using-sage}). This was added because Sage becomes
more and more the standard open-source CAS system. Accordingly all
samples written before in PARI-GP and Mathematica have been substituted
with Sage code. Thanks to Minh Van Nguyen a lot of new code samples could be added.

@@

I hope that many readers have fun with this script and that they get 
out of it more interest and greater understanding of this modern but 
also very ancient topic.
\\
\\
% [1.5\baselineskip]
% \enlargethispage*{2\baselineskip}
% \nopagebreak
Bernhard Esslinger
\\
\\
% [\baselineskip]
Frankfurt (Germany), July 2009
\vspace{-12pt}

@@

At this point I'd like to thank explicitly the following people who
particularly contributed to CrypTool\index{CrypTool}. They applied their
very special talents and showed really great engagement:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Staff of Prof.\ Claudia Eckert, Prof.\ Johannes Buchmann and Prof.\ Torben Weis.
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\
\\
Bernhard Esslinger
\\
\\
Frankfurt (Germany), July 2009

### 2009-07-07 script/en/chapters/numbertheory.tex
CT script: Some small updates and corrections done in German and English

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice)\index{Mathematica}:
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\index{Pari-GP}\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or \index{BC} BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\index{Mathematica}\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.\index{Pari-GP}
\vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +6 pt
} is extremely difficult to solve.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_12}{}
\section{The RSA procedure with actual numbers}
\label{rsaconcrete}\index{RSA}
Having described above \hyperlink{RSA}{how the RSA procedure works}, we will now work through the steps using actual, but small, numbers.

@@

\bibitem[Agrawal2002]{Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002 \\
    \href{http://www.cse.iitk.ac.in/news/primality.html}
    {\texttt{http://www.cse.iitk.ac.in/news/primality.html}}

>>>>

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice)\index{Mathematica}:
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\index{Pari-GP}\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or \index{BC} BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\index{Mathematica}\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.\index{Pari-GP}
\vskip +6 pt  % \vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +18 pt
} is extremely difficult to solve.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_12}{}
\section{The RSA procedure with actual numbers\footnotemark}
\footnotetext{%
    \index{Sage}%
    \index{Minh Van Nguyen}%
    Additional material: Minh Van Nguyen: "Number Theory and the RSA Public Key
    Cryptosystem. Introductory tutorial on using Sage to study elementary number theory
    and public key cryptography", 2009.
    Didactically very clear article about some basics in NT and in Sage usage.
    \href{http://nguyenminh2.googlepages.com/sage_numtheory-rsa.pdf}
       {\texttt{http://nguyenminh2.googlepages.com/sage\_numtheory-rsa.pdf}}.
    \vskip +3 pt
}
\label{rsaconcrete}\index{RSA}
Having described above \hyperlink{RSA}{how the RSA procedure works}, we will now work through the steps using actual, but small, numbers.

@@

\bibitem[Agrawal2002]{Agrawal2002}  \index{Agrawal 2002} 
    M. Agrawal, N. Kayal, N. Saxena, \\
    {\em PRIMES in P}, August 2002, revised paper: \\
       \href{http://www.cse.iitk.ac.in/~manindra/algebra/primality_v6.pdf}
    {\texttt{http://www.cse.iitk.ac.in/~manindra/algebra/primality\_v6.pdf}} \\
    See also the website "The AKS "PRIMES in P" Algorithm Resource":
       \href{http://fatphil.org/maths/AKS/}
    {\texttt{http://fatphil.org/maths/AKS/}}
